---
title: "happiness_analysis"
output:
  pdf_document: 
    
    latex_engine: xelatex
  html_document:
    df_print: paged
geometry: top=0.5cm
date: "2024-12-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#### Partial residual plots ####

ls.part.res.plot <- function(x,y,j,intercept=TRUE,x.name=NULL,y.name="Y",main=NULL)
  
  # Produces a partial-residual plot for variable j in x. 
  # x - X matrix for current model (no column of 1's should be included for
  #     for an intercept term);
  # y - Y vector;
  # j - a column number of x;
  # intercept - set to FALSE for a no-intercept model;
  # x.name - x-axis label - if not supplied, then the name of column
  #          j in x will be used;
  # y.name - y-axis label;
  # main - main title for plot, if not supplied then the title is of the form
#        "Partial-residual plot for x".

{ # Get residuals. 
  reg <- lsfit(x,y,intercept=intercept)
  res <- reg$residuals
  
  
  # add component from x_j
  # if there is an intercept, we want coeeficient j+1
  
  jj <- j
  if(intercept == TRUE)
    jj <- j+1
  res <- res+reg$coef[jj]*x[,j]
  
  # set up labels. 
  
  if(mode(x.name) == "NULL")
    x.name <- dimnames(x)[[2]][j]
  if(mode(main) == "NULL")
    main <- paste("Partial-residual plot for", x.name)
  
  
  plot(x[,j], res, main=main, xlab=x.name,
       ylab=paste("Partial residuals for",y.name))
  lines(lowess(x[,j],res))
}

##### Added-variable plot #####

ls.added.var.plot <- function(x,y,added.var,intercept=TRUE,
                              x.name="X",y.name="Y",main="")
  # Produces an added-variable plot.
  # x - X matrix for current model (no column of 1's should be included for
  #     for an intercept term);
  # y - Y vector;
  # added.var - if a scalar, a column number of x;
  #              if a vector of length > 1, values for a "new" x variable.
  # intercept - set to FALSE for a no-intercept model;
  # x.name - x-axis label - if added.var is a scalar, the default "x" is 
  #           replaced by the name of column added.var;
  # y.name - y-axis label;
  # main - main title for plot, if not supplied then the title is of the form
#        "Added-variable plot for x".
{
  if(length(added.var)==1)
  { 
    # added.var is a column number for a variable in x.
    if(x.name =="X")
      x.name <- dimnames(x)[[2]][added.var]
    xtilde <- x[,-added.var]
    s.res <- lsfit(xtilde,y,intercept=intercept)$residuals
    t.res <- lsfit(xtilde,x[,added.var],intercept=intercept)$residuals
  }
  else
  {
    # added.var is a "new" x variable not in x.
    s.res <- lsfit(x,y,intercept=intercept)$residuals
    t.res <- lsfit(x,added.var,intercept=intercept)$residuals
  }
  
  if(main =="")
    main <- paste("Added-variable plot for", x.name)
  
  plot(t.res,s.res,main=main,xlab=paste("Corrected",x.name),
       ylab=paste("Corrected",y.name))
  lines(lowess(t.res,s.res))
}

```


We will be performing residual analysis on 2019_happiness_cleaned.csv, the Happiness dataset for 2019.

Installing packages
```{r}
#Install and import package to deal with the categorical variable, region.
if (!require(fastDummies)) install.packages('fastDummies', dependencies = TRUE)
library(fastDummies)
```

Now we read in the CSV file.

```{r}
#Import the csv file
HAPPY <- read.csv("2019_happiness_cleaned.csv")
# Remove rank, as we will be using score as the response variable
HAPPY <- subset(HAPPY, select = -c(rank))
head(HAPPY)
```

```{r}
#Create dummy variables
HAPPY2 <- dummy_cols(HAPPY, select_columns = "region")
# Drop the `region`, because we have converted it into dummy variables
# Drop the`Australia and New Zealand` region, because we need to ensure t(X) %*% X will be non-singular.
HAPPY2 <- subset(HAPPY2, select = -c(region, `region_Australia and New Zealand`))
head(HAPPY2)
```

We will define the X matrix. This matrix will consist of these columns, in this order:
Intercept, GDP per Capita, Social support, Healthy life expectancy, The freedom to make life choices, Generosity, Perceptions of corruption, and all the Region variables (except the Australia and New Zealand region).


```{r}
X <- cbind(1, as.matrix( HAPPY2[,c(2:ncol(HAPPY2))] ))
head(X)
```
We will define the Y vector. This vector will contain the happiness scores.

```{r}
Y <- as.matrix( HAPPY2[,c(1)] )
head(Y)
```
```{r}
LMHappy = lm(score~., data=HAPPY2)
summary(LMHappy)
```
```{r}
anova(LMHappy)
```


```{r}
plot(X[,2], Y, xlab="GDP", ylab="Happiness Score")
plot(X[,3], Y, xlab="Support", ylab="Happiness Score")
plot(X[,4], Y, xlab="Life Expectancy", ylab="Happiness Score")
plot(X[,5], Y, xlab="Freedom", ylab="Happiness Score")
plot(X[,6], Y, xlab="Generosity", ylab="Happiness Score")
plot(X[,7], Y, xlab="Corruption", ylab="Happiness Score")
```


1. Residual plot against each explanatory variable,
2. Partial residual plot for each explanatory variable,
3. Added variable plot.

4. Residuals vs. fitted,
5. Q-Q residuals (theoretical quantities vs. standardized residuals).

```{r}
happy.lm <- lm(score~.,data=HAPPY2)

happy.res.before <- happy.lm$residuals

par(mfrow=c(2,3))

plot(X[,2], happy.res.before, xlab="GDP", ylab="residuals", main="GDP Residual Plot")
ls.part.res.plot(X[,2:ncol(X)],Y,1)
ls.added.var.plot(X[,2:ncol(X)],Y,1)

plot(X[,3], happy.res.before, xlab="Support", ylab="residuals", main="Support Residual Plot")
ls.part.res.plot(X[,2:ncol(X)],Y,2)
ls.added.var.plot(X[,2:ncol(X)],Y,2)

plot(X[,4], happy.res.before, xlab="Life Expectancy", ylab="residuals",main="Life Expectancy Residual Plot")
ls.part.res.plot(X[,2:ncol(X)],Y,3)
ls.added.var.plot(X[,2:ncol(X)],Y,3)

plot(X[,5], happy.res.before, xlab="Freedom", ylab="residuals", main="Freedom Residual Plot")
ls.part.res.plot(X[,2:ncol(X)],Y,4)
ls.added.var.plot(X[,2:ncol(X)],Y,4)

plot(X[,6], happy.res.before, xlab="Generosity", ylab="residuals", main="Generosity Residual Plot")
ls.part.res.plot(X[,2:ncol(X)],Y,5)
ls.added.var.plot(X[,2:ncol(X)],Y,5)

plot(X[,7], happy.res.before, xlab="Corruption", ylab="residuals", main="Corruption Residual Plot")
ls.part.res.plot(X[,2:ncol(X)],Y,6)
ls.added.var.plot(X[,2:ncol(X)],Y,6)
```


```{r}
fitB <- solve(t(X) %*% X) %*% t(X) %*% Y
fitY <- fitted(lm(score~., data=HAPPY2))
p = ncol(X) - 1
n = length(Y)

SSreg = t(fitY) %*% fitY - n * mean(Y)^2
DFreg = p
MSreg = SSreg/DFreg
cbind(SSreg, DFreg, MSreg)

SSres = t(Y - fitY) %*% (Y - fitY) 
DFres = n-p-1
MSres = SSres/DFres
cbind(SSres, DFres, MSres)

SStot = SSreg+SSres
DFtot = n-1
cbind(SStot, DFtot)

```

```{r}

```